---
title: "Conservation Analysis with TrajConserve"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conservation Analysis with TrajConserve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Conservation Analysis with TrajConserve

This vignette focuses on the conservation analysis capabilities of the TrajConserve package. We'll explore how to identify conserved gene expression patterns across samples, batch conditions, or experimental replicates.

## What is Conservation?

In the context of TrajConserve, "conservation" refers to the consistency of gene expression patterns across different conditions. A gene is considered conserved if:

1. It has high expression values (high mean estimate)
2. It shows consistent expression across samples (low variability)

Genes that have low expression or high variability across samples are considered non-conserved.

## The Conservation Score 

TrajConserve quantifies conservation using a weighted score that combines two key aspects:

1. **Mean Expression**: Higher mean expression contributes to a higher conservation score
2. **Expression Variability**: Lower variability (measured by coefficient of variation) contributes to a higher conservation score

The relative importance of these factors can be adjusted using weight parameters.

## Computing Conservation Scores

### Basic Usage

```{r, eval=FALSE}
library(TrajConserve)

# Assuming you have already run models and saved results to an HDF5 file
h5_file <- "trajectory_models.h5"

# Calculate conservation scores with default parameters
conservation_results <- calculate_conservation(
  h5_file = h5_file,
  metric = "Estimate"
)

# View the results
head(conservation_results)
```

### Customizing the Analysis

You can adjust several parameters to customize the conservation analysis:

```{r, eval=FALSE}
# Customize the weights and threshold
conservation_results <- calculate_conservation(
  h5_file = h5_file,
  metric = "Estimate",
  mean_weight = 0.7,           # Give more importance to mean expression
  variability_weight = 0.3,    # Give less importance to variability
  conservation_threshold = 0.8, # Stricter threshold for conserved genes
  normalize_scores = TRUE      # Normalize scores to 0-1 range
)
```

## Understanding the Results

The `calculate_conservation` function returns a data frame with the following columns:

- `gene`: Gene identifier
- `mean_estimate`: Mean expression value
- `sd_estimate`: Standard deviation of expression
- `cv`: Coefficient of variation (sd/mean)
- `range_estimate`: Range of expression values
- `mean_norm`: Normalized mean value (if normalize_scores = TRUE)
- `cv_norm`: Normalized inverse variability
- `conservation_score`: Overall conservation score
- `is_conserved`: Boolean indicating if the gene is conserved (based on threshold)

## Visualizing Conservation

### Scatter Plot

The scatter plot shows the relationship between mean expression and variability:

```{r, eval=FALSE}
# Create a scatter plot
plot_conservation(
  conservation_results,
  plot_type = "scatter",
  highlight_n = 10,  # Highlight top/bottom 10 genes
  file_path = "conservation_scatter.pdf"  # Optional: save to file
)
```

### Histogram

The histogram shows the distribution of conservation scores:

```{r, eval=FALSE}
# Create a histogram
plot_conservation(
  conservation_results,
  plot_type = "histogram",
  file_path = "conservation_histogram.pdf"
)
```

### Heatmap

For creating heatmaps, you'll need the original expression data:

```{r, eval=FALSE}
# Extract the data
estimate_matrix <- extract_hdf5_metric(h5_file, "Estimate")

# Create a heatmap
# Note: Using pheatmap directly is recommended for complex heatmaps
library(pheatmap)

# Create annotation for conserved vs non-conserved genes
gene_type <- ifelse(conservation_results$is_conserved, "Conserved", "Non-conserved")
gene_anno <- data.frame(
  Conservation = factor(gene_type, levels = c("Conserved", "Non-conserved")),
  row.names = conservation_results$gene
)

# Define colors for annotation
anno_colors <- list(
  Conservation = c(Conserved = "blue", `Non-conserved` = "red")
)

# Create heatmap
pheatmap(t(estimate_matrix), 
         annotation_row = gene_anno,
         annotation_colors = anno_colors,
         main = "Expression Patterns of Conserved vs Non-conserved Genes",
         cluster_rows = TRUE,
         cluster_cols = TRUE)
```

## Comparing Conservation Across Different Metrics

You can calculate conservation for different metrics in your HDF5 file:

```{r, eval=FALSE}
# Compare Estimate and Est.Error metrics
estimate_conservation <- calculate_conservation(h5_file, "Estimate")
error_conservation <- calculate_conservation(h5_file, "Est.Error")

# Compare the number of conserved genes
table(estimate_conservation$is_conserved)
table(error_conservation$is_conserved)

# Find genes conserved in both metrics
both_conserved <- intersect(
  estimate_conservation$gene[estimate_conservation$is_conserved],
  error_conservation$gene[error_conservation$is_conserved]
)
print(both_conserved)
```

## Case Study: Finding Conserved Genes in Development

Let's walk through a complete example of finding conserved genes in a developmental trajectory:

```{r, eval=FALSE}
library(TrajConserve)
library(dplyr)

# 1. Load results from HDF5 file
h5_file <- "development_trajectory_models.h5"

# 2. Calculate conservation scores
cons_results <- calculate_conservation(
  h5_file = h5_file,
  mean_weight = 0.6,
  variability_weight = 0.4,
  conservation_threshold = 0.7
)

# 3. Analyze results
# Number of conserved and non-conserved genes
table(cons_results$is_conserved)

# Top 10 most conserved genes
top_conserved <- cons_results %>%
  filter(is_conserved) %>%
  arrange(desc(conservation_score)) %>%
  head(10)

print(top_conserved[, c("gene", "conservation_score", "mean_estimate", "cv")])

# Bottom 10 least conserved genes
least_conserved <- cons_results %>%
  arrange(conservation_score) %>%
  head(10)

print(least_conserved[, c("gene", "conservation_score", "mean_estimate", "cv")])

# 4. Visualize results
# Scatter plot
plot_conservation(cons_results, plot_type = "scatter", highlight_n = 15)

# Extract data for heatmap
data_matrix <- extract_hdf5_metric(h5_file, "Estimate")

# Get top 20 most conserved and 20 least conserved genes
genes_to_plot <- c(
  cons_results$gene[order(-cons_results$conservation_score)][1:20],
  cons_results$gene[order(cons_results$conservation_score)][1:20]
)

# Subset data
plot_data <- data_matrix[, genes_to_plot]

# Create annotation data frame
gene_type <- rep("Non-conserved", length(genes_to_plot))
gene_type[genes_to_plot %in% cons_results$gene[cons_results$is_conserved]] <- "Conserved"

gene_anno <- data.frame(
  Conservation = factor(gene_type, levels = c("Conserved", "Non-conserved")),
  row.names = genes_to_plot
)

# Plot heatmap
pheatmap(t(plot_data),
         annotation_row = gene_anno,
         scale = "row",
         main = "Expression Patterns Across Samples",
         fontsize_row = 8)
```

## Biological Interpretation

When interpreting conservation results:

- Highly conserved genes likely represent core biological processes that are stable across conditions
- Non-conserved genes may indicate:
  - Condition-specific responses
  - Technical variability
  - Genes undergoing dynamic changes
  
- Consider gene functions and pathways when interpreting conservation patterns:
  - Housekeeping genes are often highly conserved
  - Genes involved in stress response, immune function, or developmental transitions may show lower conservation

## Conclusion

Conservation analysis is a powerful tool for identifying genes with consistent expression patterns across experimental conditions. By quantifying both expression level and variability, TrajConserve provides a robust approach to finding biologically meaningful patterns in your trajectory data. 