---
title: "Introduction to TrajConserve"
author: "Gilbert Han"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to TrajConserve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# TrajConserve

## Overview

TrajConserve is an R package designed for analyzing gene expression trajectories and identifying conserved expression patterns across experimental conditions. This package implements Bayesian generalized additive models (GAMs) to model gene expression over pseudotime trajectories and provides tools for quantifying conservation of expression patterns.

The package is particularly useful for:

- Modeling expression dynamics along developmental or cell differentiation trajectories
- Identifying genes with conserved or divergent expression patterns across conditions
- Visualizing and comparing trajectory models
- Storing and retrieving model results in HDF5 format

## Features

- **Trajectory Modeling**: Fit Bayesian GAMs to gene expression data along pseudotime trajectories.
- **Conservation Analysis**: Quantify gene conservation based on expression levels and variability.
- **HDF5 Integration**: Store and retrieve large model results efficiently.
- **Visualization Tools**: Create publication-ready plots of model results and conservation metrics.

## Installation

You can install TrajConserve from GitHub:

```{r, eval=FALSE}
# Install devtools if you haven't already
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")

# Install the package from GitHub
devtools::install_github("username/TrajConserve")
```

### Dependencies

TrajConserve depends on several packages, including:

- **brms**: For Bayesian regression modeling via Stan
- **rhdf5**: For HDF5 file handling
- **ggplot2**: For visualization
- **pheatmap**: For creating heatmaps
- **data.table**: For efficient data manipulation

## Basic Usage

### 1. Converting Seurat data to trajectory array

```{r, eval=FALSE}
library(TrajConserve)
library(Seurat)

# Assuming 'seurat_obj' is your Seurat object with integrated data
# and 'pseudotime' is a column in the metadata
trajectory_data <- seurat_to_trajectory_array(
  seurat_obj = seurat_obj,
  assay = "RNA",
  pseudotime_col = "pseudotime",
  batch_col = "sample_id",
  genes = c("Gene1", "Gene2", "Gene3", "Gene4")
)
```

### 2. Running trajectory models

```{r, eval=FALSE}
# Run models for multiple genes
result <- run_multiple_models(
  data_array = trajectory_data,
  output_file = "trajectory_models.h5",
  family = "negbinomial",
  n_cores = 4
)
```

### 3. Analyzing conservation of gene expression

```{r, eval=FALSE}
# Calculate conservation scores
conservation_results <- calculate_conservation(
  h5_file = "trajectory_models.h5",
  metric = "Estimate",
  mean_weight = 0.6,
  variability_weight = 0.4,
  conservation_threshold = 0.7
)

# View conservation results
head(conservation_results)

# Visualize conservation metrics
plot_conservation(conservation_results, plot_type = "scatter")
```

### 4. Extracting and visualizing metrics from HDF5 files

```{r, eval=FALSE}
# Extract Estimate values
estimates <- extract_hdf5_metric("trajectory_models.h5", "Estimate")

# Create a heatmap
plot_hdf5_heatmap("trajectory_models.h5", "Estimate", scale = "row")
```

## Example Workflow

Below is a complete workflow example from preprocessing data to analyzing conservation:

```{r, eval=FALSE}
library(TrajConserve)
library(Seurat)

# 1. Prepare data
seurat_obj <- readRDS("your_seurat_object.rds")
genes_of_interest <- c("Gene1", "Gene2", "Gene3", "Gene4")

# 2. Convert Seurat data to trajectory array
traj_array <- seurat_to_trajectory_array(
  seurat_obj = seurat_obj,
  assay = "RNA",
  pseudotime_col = "pseudotime",
  batch_col = "condition",
  genes = genes_of_interest
)

# 3. Run trajectory models
output_file <- "results/trajectory_models.h5"
run_multiple_models(
  data_array = traj_array,
  output_file = output_file,
  family = "negbinomial",
  n_cores = 4
)

# 4. Calculate conservation scores
conservation_results <- calculate_conservation(
  h5_file = output_file,
  metric = "Estimate",
  mean_weight = 0.6,
  variability_weight = 0.4,
  conservation_threshold = 0.7
)

# 5. Create output directory
output_dir <- "results/figures"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Create a heatmap with gene conservation annotation
library(pheatmap)

# Extract estimate matrix
estimate_matrix <- extract_hdf5_metric( output_file, "Estimate")

# Create annotation data frame
gene_type <- ifelse(conservation_results$is_conserved, "Conserved", "Non-conserved")
gene_anno <- data.frame(
  Conservation = factor(gene_type, levels = c("Conserved", "Non-conserved")),
  row.names = conservation_results$gene
)

# Create heatmap
pheatmap(t(estimate_matrix), 
         annotation_row = gene_anno,
         main = "Expression Patterns of Conserved vs Non-conserved Genes")

# 7. Print top conserved genes
top_conserved <- conservation_results[order(-conservation_results$conservation_score), ][1:10, ]
print(top_conserved[, c("gene", "conservation_score", "is_conserved", "mean_estimate", "cv")])
```

## Advanced Topics

For more advanced usage, please see the other vignettes:

- [workflow](workflow.html)
- [Statistical background](https://scatlas.readthedocs.io/en/latest/trajectory/20250330_statistical_trajconserve.html)
- [Conservation analysis in depth](conservation_analysis.html)